# Tools - Creative use of audio effects

```{r}
#| echo: FALSE
knitr::opts_chunk$set(echo = FALSE)
```

It ***is possible*** to create generative tracks that do not use reverb and delay. But to be honest, it isn't done often. In the preface to this book I said about ambient music: "Take any sound, drown it in reverb and delay... JOB DONE". You can, and I'd argue you ***should*** do better than that. Creative use of audio effects can accentuate harmonics and do interesting things with those; provide some grit, noise or warble; take a rather plain sound and use modulation and effects to twist it until it forms something much more interesting and evolving. All this BEFORE drowning it in reverb and delay.

::: callout-tip
## Key idea

Audio effects and automation together can take sounds and keep them interesting and evolving in ways that go beyond "static" application of reverb and delay on Return channels. Think of how you can use the huge range of creative effects provided by Ableton and employ these in ways that will take your original sound in completely new directions. This will provide the listener with some additional "ear candy" to keep the sounds interesting.

::: 

## The tools of the trade

Ableton Live Suite comes with a HUGE range of audio effects and if you add in Max for Live effects there are almost too many audio-manglers out there to list or talk about in this chapter. While there are a few third-party audio effect plugins that are worth a mention, and which I use a LOT in my productions, you can get a LONG way with just the stock Ableton Live Suite audio effects. The benefit of third party plugins is where they offer something that the Live effects cannot. I'll talk about some of these below. But let's start looking at a few key plugins which are likely to be useful for generative music (and in particular for generative, ambient music which I create).

## Hybrid Reverb

The [Hybrid Reverb](https://www.ableton.com/en/live-manual/11/live-audio-effect-reference/#hybrid-reverb) in Live Suite is a really nice reverb. It has a Convolution Impulse Response (IR) section which allows you to dial up real-world reverb from a wide variety of spaces and an Algorithm section which applies reverb processing and colouration. You can blend between these or apply them in parallel and series. This in itself offers up a huge variety of sounds. This video will help explore what can be done with the Hybrid Reverb effect <https://www.youtube.com/watch?v=yLBIOiM97Vs>.

But when you dig into the Convolution IR section you'll also find options for textural processing - this applies a textural Impulse Response (IR) to the sound, so instead of a straight reverb tail, you'll get some additional texture from the IR audio file. The other thing to point out is that there is an option to have User-defined IR. This means that you can find IR audio files for different spaces online at sites like Open Air <https://www.openair.hosted.york.ac.uk> and use some additional IR files for some famous spaces, like Hamilton Mausoleum and York Minster. Obviously finding the right reverb sound for ambient music is kind of a key item. But you don't have to use just one. I like to use a long, lush reverb to help smear sound and create drones. I usually put this on a FX Return channel so I can route several tracks to it and save CPU. But you may also need a shorter reverb for other instruments to give them presence, in the way that a close mic to an instrument in one of these huge reverb spaces will capture more of the raw sound, while the gigantic reverb captures the instrument in the space itself.

```{r}
#| fig-alt: "The image shows the Hybrid Reverb audio device. The device is dominated by the middle panel which shows the impulse response for the chosen Convolution IR reverb. There are selectors for the type of reverb and space for the IR section, also above that options to tune the attack, decay and size of the IR reverb. On the bottom right are settings for the algorithmic part of the reverb with dials for Decay, Size, Damping, and two other parameters which are context specific to the algorithm chosen. Here Shimmer reverb is chosen so Pitch and Shimmer amount are showing. A middle dial allows you to balance between the IR and Algorithm reverb and this is set to 50/50. To the right is a dial to adjust stereo image, a button for Vintage mode, a button to select Bass in Mono and a Dry/Wet dial."
knitr::include_graphics(path = file.path(here::here(), "images", "Hybrid_Reverb.png"))
```

## Valhalla Supermassive

Here's a third-party effect I couldn't do without. [Valhalla Supermassive](https://valhalladsp.com/shop/reverb/valhalla-supermassive/) is a free audio effect plugin from Valhalla DSP. It's a staple of ambient and drone producers because a) it's free and b) it has some MASSIVE reverb algorithms (duh!). Suffice to say it does exactly what it says on the tin - massive reverbs. But what Valhalla have done rather well is to combine algorithms and delay lines so that some of its algorithms behave more like delays, while some act more like reverbs. They also have algorithms that combine both - early delay turning into massive reverb. Some have short attack, some much longer. One of my personal favourites is the "Benson Arizona" preset which will turn ANYTHING into a massive drone. Combine this with Paul Stretch, as discussed in 
[Tools - Extreme audio stretching with PaulStretch](016-Tools-PaulStretch.qmd)
and you'll have instant drones.

To be honest, all of Valhalla DSP's plugins are excellent and reasonably priced. I particularly like the Shimmer device to add shimmer, stereo width and pitch shift reverb to sounds. Their Delay plugin provides a wide variety of delay types, including tape delay (complete with wobble and drive) but also usefully a reverse delay (where the original audio is played backwards in the delay). I like the latter as the delayed line comes back transformed compared to the input. Again this adds some interest to the part.

```{r}
#| fig-alt: "The image shows the Valhalla Supermassive device from Valhalla DSP. To the left are dials to adjust Mix and Width. Then there are two large dials controlling Delay and Warp. To the right of that are two vertically stacked dials controlling Feedback and Density. Another pair of stacked dials controls the Mod Rate and Depth. The final pair of stacked dials controls EQ settings. There are dropdown picklists below with Mode on the bottom left and Preset on the bottom right."
knitr::include_graphics(path = file.path(here::here(), "images", "Valhalla_Supermassive.png"))
```

## Echo

I tend to prefer [Live Suite's Echo effect](https://www.ableton.com/en/live-manual/12/live-audio-effect-reference/#echo) rather than stock delay. There is a wide range of sounds possible using it, and some built in modulations and tweaks to the sound which add character. Use of Echo as a send effect has a long legacy in dub music where the DJ or producer would send a burst of signal to the Echo and by tweaking the feedback and EQ would create long tails of echo. You can learn more about the Echo effect here: <https://www.youtube.com/watch?v=4LxhIE169x4>.

Being a Lazy Producer, I prefer to let random processes tweak the Echo settings. I have built an audio effect rack with a trio of LFOs working on each other and then changing the feedback amount, lower and upper ranges of the EQ so that these wobble about randomly. Using an instance of Dillon Bastan's Strange Mod modulator will achieve similar effect as discussed in 
[Tools - Modulators](007-Tools-Modulators.qmd)
 While the timing of the Echo is fixed, we have other parameters that mean that the sound is constantly shifting in an unpredictable way. This will prevent the Echo effect from being too "samey" across the track and provide some interesting moments. It's a bit like having a pair of "robot hands" that tweak the Echo settings.

```{r}
#| fig-alt: "The image shows the Ableton Echo device. The left of the device sets the delay characteristics for both left and right channels. Delay times can be synced to tempo divisions as they are here, or can be set in Hz. The left and right delay lines can be locked together. Below these is a dial for Input gain and Feedback. The middle of the device shows a graphical representation of the delay with concentration half-circles representing the echos. Below this is an filter representation for high pass and low pass filters. To the right of the device are dials for Reverb amount and stacked dials for Stereo width, Output and Dry Wet. There are buttons to set the Echo mode including Stereo, Ping Pong and Mid/Side. Other tabs control the modulation settings and character of the Echo. To the right of the Echo device are three LFOs set at various rates. The left one controls the Echo Feedback amount. The other LFOs are controlling the Depth and Rate of the first LFO. The first LFO also controls the Depth of the third LFO."
knitr::include_graphics(path = file.path(here::here(), "images", "LFO_Echo.png"))
```

## Saturation and Distortion

Saturation may seem like a weird choice for ambient music where much of the music produced is "pristine" and "delicate". But there is a whole sub-genre where gritty, distorted sounds are most definitely a thing. My favourite from Live Suite is the [Saturator plugin](https://www.ableton.com/en/live-manual/11/live-audio-effect-reference/#saturator) which provides a wide range of subtle distortion and drive. If you need more distortion than saturation then the [Pedal](https://www.ableton.com/en/live-manual/11/live-audio-effect-reference/#pedal) plugin provides raw Overdrive, Distortion and Fuzz, Saturator allows you to drive the signal and dial in more subtle drive and overtones.

```{r}
#| fig-alt: "The image shows the Saturator device. On the top left is a Drive dial, below that buttons for DC offset, Color and a dial for Base frequency. In the middle is a graphical representation of the gain curve and how saturation is going to affect input gain. Below that is a dropdown box for the saturator type. Three dials below that adjust Frequency, Width and Depth. To the right are dials for Output gain and Dry / Wet setting. Next to the Saturator device is a Pedal device which has dials on the top row adjusting Gain percentage and Output in dB. Below that is a row of buttons to set the distortion type: Overdrive (marked OD), Distortion (marked Distort) and Fuzz. Below that are three dials adjusting Bass, Mid and Treble percentage. On the bottom row is a button for turning on Sub frequency and a Dry/Wet slider."
knitr::include_graphics(path = file.path(here::here(), "images", "Saturator_Pedal.png"))
```

Live 12's [Roar](https://www.ableton.com/en/live-manual/12/live-audio-effect-reference/#roar) device takes these to the next level though and provides a comprehensive set of tools for applying anything from subtle warmth to the inputs to full-blown distortion. It has a wide range of routing options, including multi-band which allows you to dial in distortion for different frequency bands. This is a device that is really worth exploring to add texture in a wide variety of ways.

```{r}
#| fig-alt: "The image shows the Roar device in Live 12. There are three main panels in the device. Starting on the left there is a Drive (in dB) and Tone (in %) dials including a slider to set the mid-point frequency for Tone. Below that is a dropdown box to pick the routing of audio through Roar. Here we're showing the Multi-band option, which has associated Low and High frequency value boxes to change the crossover. The middle panel depends on the Roar type setting, so here it shows tabs for Low, Mid and High frequency saturation. Each tab shows dials for Amount (in %), Bias (-/+) and Frequency (in Hz). Below the dials are two graphical representations: on the left is the saturation curve and on the right is the filter, which here is a Low pass filter. To the right is a panel devoted to Feedback mode with a dial for amount. To the right of that is a compression dial (in %) and an output dial (gain dB) and at the bottom right a Dry/Wet % slider."
knitr::include_graphics(path = file.path(here::here(), "images", "Roar.png"))
```

More about Roar here: <https://www.youtube.com/watch?v=ETzf6O9-6us>

## Resonators

The [Resonator effect](https://www.ableton.com/en/live-manual/11/live-audio-effect-reference/#resonators) takes an incoming signal and then resonates it by pitch-shifting additional resonators. You can dial-in the amount that the resonant frequencies are heard via gain adjustment, you can define the pitch of the resonator and whether the resonator is centred on a particular pitch. This can be useful to create drone noises and coupled with a long reverb like the Valhalla Supermassive you can create some really lush ambient beds.

I have found that a little drive to the input signal (which boosts upper resonant frequencies of the audio) works will with resonator to make the result more pronounced.

```{r}
#| fig-alt: "The image shows the Resonator device. The left hand panel has Filter settings including a button switch on and off, Frequency dial (in Hz) and type (low pass, high pass, band pass and band cut). In the middle panel is the resonance structure. On the left is a stack of dials for Decay and Color. To the right of that is a stack of dials for setting the root Note pitch and Gain, then Stacked dials for Pitch and Gain for four additional resonators. The top dial is the relative pitch in semitones, below that is a slider for panning and below that the Gain (in dB). To the far right are stacked dials for Width (in %), Gain (in dB) and Dry/Wet (in %)."
knitr::include_graphics(path = file.path(here::here(), "images", "Resonator.png"))
```

More about the Resonator effect here: <https://www.youtube.com/watch?v=wSjp6nX3rYI>

## Spectral processing

Ableton Live Suite includes three Spectral audio effects. Spectral processing is interesting because instead of applying effects in a time-based manner (as audio comes through the effect) it is applying effects according to frequency (low to high) and loudness.

### Ableton Spectral effects

[Spectral Time](https://www.ableton.com/en/live-manual/11/live-audio-effect-reference/#spectral-time) which can be either apply a freezing algorithm to catch and hold incoming audio, or delay which applies delay and frequency shifting. The Tilt, Spray and Mask parameters apply the delayed audio to different parts of the frequency spectrum and gives some interesting smeared, granular and glitchy sounds. The Freezer and Delay algorithms can be used separately or chained serially from one into the other. Setting long fade in and fade out times allows you to catch and smear incoming audio which may be useful as part of a performance to join between two tracks or two sections of a mix.

[Spectral Resonator](https://www.ableton.com/en/live-manual/11/live-audio-effect-reference/#spectral-resonator) is a resonator which focuses on different parts of the frequency spectrum, so you can tailor the resonance quite precisely. You can have a static tone for the resonance or pass in a MIDI part to change the resonance around a MIDI part. The different resonance algorithms - Chorus, Wander and Granular all have different sounds and it's worth playing with these to see which ones fit what you're trying to achieve.

```{r}
#| fig-alt: "The image shows three Spectral devices - Spectral Blur, Spectral Resonator and Spectral Time. Spectral Blur has three dials to set the lower and upper Frequencies (Freq1 and Freq2 respectively) and a Residual dial (in %). Halo and Freeze buttons immediately to the right of that with a Halo (%) dial to the right of the buttons. Below the Halo dial is a Wet Gain slider (in dB) and a Dry/Wet dial (in %). There is a graphical representation of what the device is doing - that is, catching frequencies and prolonging their resonance."
knitr::include_graphics(path = file.path(here::here(), "images", "Spectral_Devices.png"))
```

Rishabh Rajan has useful YouTube videos explaining the Spectral Time (<https://www.youtube.com/watch?v=KUZLXAK8do4>) and Spectral Resonator (<https://www.youtube.com/watch?v=VAebi_brjEs>) effects.

### Michael Norris Soundmagic Spectral collection

Michael Norris has created some fantastic spectral plugins which can be downloaded here: <https://www.michaelnorris.info/software/soundmagic-spectral> as donation-ware.

The Spectral Averaging and Spectral Blurring are my two favourites from this collection to smear and blur incoming audio and create lush drones with the same sonic "fingerprint" as the original, but as a long drone texture. Applying these to the an audio clip or to a PaulStretch stretched audio will apply even more smearing and blurring to create fantastic drone textures.

The key parameter in both of these plugins is the FFT size which determines the size of the sample window in which the blurring or averaging is taking place. Larger FFT sizes smear more, but take longer to fade in and out since the processing is happening over a larger window of audio. You may also find that you need to increase the gain of the output.

```{r}
#| fig-alt: "The image shows Michael Norris' Spectral Averaging Max for Live device. The device is very bare-bones in interface with four sliders changing the Number of frames, Brightness, Lo bin cutoff, Hi bin cutoff. Below this is the key parameter FFT Size which defines the window over which the averaging will happen. The bottom slider is for Gain."
knitr::include_graphics(path = file.path(here::here(), "images", "Spectral_Averaging.png"))
```

```{r}
#| fig-alt: "The image shows Michael Norris' Spectral Blurring Max for Live device. Like the Spectral Averaging device this is very bare bones interface. A slider at the top influences Blur amount. Below that is a Variance type dropdown. Beneath that are sliders for Blur variance, LFO rate, Lo bin cutoff and Hi bin cutoff. Like the Averaging plugin there is an FFT Size dropdown adn Gain slider."
knitr::include_graphics(path = file.path(here::here(), "images", "Spectral_Blurring.png"))
```

The whole collection is worth checking out for some very interesting effects, although as you can see above the user interface is VERY bare-bones.

## Other notable mentions

### Shifter

[Shifter](https://www.ableton.com/en/live-manual/11/live-audio-effect-reference/#shifter) is a combination of a pitch and frequency shifter and ring modulator. It's a bit of a beast because it can be either subtle and also extremely weird depending on how much shifting you do (and what type). This video provides a nice overview of the different modes and features: <https://www.youtube.com/watch?v=uqY8K8otbp0>

```{r}
#| fig-alt: "The image shows the Shifter device which is a ring modulator and pitch shifter. On the top left of the device are dials for Coarse and Spread. Coarse is measured in semitones while Spread is measured in cents. A button below the Spread dial marked Wide widens the stereo image. Sliders below these dials show Window in miliseconds and Tone in Hz. There is a delay button and a slider for note division of the delay and a Feedback dial next to that. Delay can be specified in note divisions or miliseconds. On the right of the device are Mode buttons for Pitch, Freq and Ring. Underneath this are a Drive button and slider for amount (in dB). Drive is greyed out unless Ring modulation mode is chosen. The central panel shows a visual representation of how the phase shift of the Left and Right signals are affected by settings. There is a slider for Duty cycle (in %) and Spin (also in %). Below this are settings for an LFO including a dropdown for shape, buttons for whether rate should be in note divisions or hertz, Rate dial (in Hz or note divisions) and Amount dial (in semitones since the image shows Pitch mode being selected)."
knitr::include_graphics(path = file.path(here::here(), "images", "Shifter.png"))
```

### Corpus

[Corpus](https://www.ableton.com/en/live-manual/11/live-audio-effect-reference/#corpus) is the audio effect companion to the Collision instrument. The input audio is used to excite a resonator - a plate, tube, membrane etc. - which then resonates. Depending on the resonator you choose you get different artefacts and results.

```{r}
#| fig-alt: "The image shows the Corpus device which is an audio effect based on the Collision instrument. While collision has an exciter section Corpus uses the incoming audio to excite the plate, membrange or other type of resonator. On the left of the device is a representation of the resonant object. In this case that is a Membrane. Attributes like Decay, Material, Width, Position Left and Right are shown. Below this are four dials for Bright, Inharmonics, Ratio and Hit (all in percent). Next is a vertical section dealing with pitch. So there is a tuning dial (in Hz but showing the MIDI note below), below this is a fine tuning dial (in cents). Below this is a Spread slider (in %). Next is a vertical LFO section with dials for Amount (in %), a dropdown box for Shape, Rate (Hz or note divisions) and Phase (in degrees). There is checkbox at the top to switch on or off this section. The next vertical section coming right is a filter section. There is a checkbox at the top to switch on or off this section. Dials below this are for Frequency (in Hz) and Bandwidth. Finally a vertical output section with three stacked dials for Bleed (in %), Gain (in dB) and Dry Wet (in %)."
knitr::include_graphics(path = file.path(here::here(), "images", "Corpus.png"))
```

## FX in series vs Effects Rack vs Return channels vs Separate tracks

There are so many routing options in Ableton Live that give us a HUGE range of possibilities when it comes to routing audio through effects. The effects can be on the track itself either in series (each effect passes its output on to the next effect) or in parallel using Effects Racks, on a separate track that brings audio from another track into it - with Monitor set to "In" and input selecting from another audio track or group bus - or in a Return channel where one or more tracks are routed into the Return send for processing. So which to choose?

I don't mind. It's YOUR track. But here are some thoughts on how ***I*** would use each option.

I would use effects in series where I want the effects to build on each other. This is helpful if you're going to use an effect to add some colour which can be exploited in the next effect, for example adding some grit that will ping a resonator, or a compressor to reduce the highs and bring up quieter moments which can then be fed into a saturator.

Effects in an Effects Rack are typically in parallel, although it's of course possible to build chains of effects within each chain or layer of the Effects Rack. I typically use these and then apply some modulation that will fade in and out each chain or select between them so that over time the different effects are not static over time.

Return channels are really useful if you'd like to send more than one track to an effect (reverb or delay) where you'd like to keep the effect consistent for each track or where you want to "mix" various tracks through an effect.

Using the output of one track as the input for another and then applying effects is a little like using a Return channel only the return channel is an ordinary Live track. This can be useful if you want to keep the "raw" track in the mix *and* apply effects separately, but it also allows you to pass the result of ***that*** track on into another track. So at any stage you can split out the audio, apply effects, route any of that signal to Return channels, then ingest the effected track into a further track for yet more processing.

So the choice is yours. Really. Try them all out. See which one works for you in a given situation. I don't mean that you should ***pick one*** and stick with it, but rather that you have a range of possible avenues to explore even within one track.

## Use an Envelope with your effect

If your sound has a sharp attack, you may want to soften the send to the FX so that only the tail of the sound gets effected. Use an [Envelope Follower](https://www.ableton.com/en/live-manual/11/max-for-live-devices/#envelope-follower) effect before the audio effect. By setting the Map to 100% lowest and 0% highest (inverting the signal) it will duck the initial part of the sound (turning the send DOWN) and then bring it back up as the audio decays. Note that in the device below I have turned up gain to accentuate the behaviour (basically to make the peak duck enough that the initial part of the audio has a low enough send amount) and I have smoothed out the Rise and Fall amounts which removes any "pumping" of the send to the Return channel effect. By setting a Delay you can also manually dial in how much or little of that initial part of the sound gets sent to the Return channel. This technique can be useful to capture and build drone sounds where you want to avoid sudden changes volume.

```{r}
#| fig-alt: "The image shows the Envelope Follower device. This device is largely taken up by a visual representation of the incoming signal which is shown using the spectrum bands, at the top of this is a modulation picker where you can select which parameter to modulate and lower and upper percentage sliders. Below the graphic are dials for Gain (in dB), Rise and Fall (in %) and Delay (in miliseconds)."
knitr::include_graphics(path = file.path(here::here(), "images", "Envelope_Follower_Send.png"))
```
