# Process - Unsynced loops {#process-unsynced-loops}

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = "100%")
```

"Around the world" - Daft Punk

Everyone loves a good loop. If you get a good idea of a pattern or something that lasts 8 bars then you can let that good thing just loop around and around and around (the world). A large part of The Lazy Producer's arsenal depends on loops and repetition - after all we're balancing predictable and unexpected, right? But if there's something in your brain right now that's saying "No, 8 bar loops repeated infinitely really ***is*** the ultimate Lazy Producing..." then I'd have to agree with you. So how do we break away from the tyranny of the 8 bar loop? Arrangements? Verse-Chorus-Middle 8 structures?

Before you abandon the ideas of generative music (which doesn't really lend itself well to arrangements, verses or choruses), there's a trick up our sleeve which is to take our loops and make one of them shorter or longer than the other.

------------------------------------------------------------------------

## Key idea {.unnumbered}

Making loops different lengths combines the predictable: each loop contains the same music idea each time - with the unexpected: the combination of those musical ideas each time the loop cycles creates something new and unheard.

------------------------------------------------------------------------

## Set phases to stunning

Steve Reich and other minimalist composers worked with musical phrases that each performer played a number of times. Sometimes the number of repetitions was at the discretion of the performer, sometimes governed by the number of times the performer could play it in one breath, sometimes the composer worked with actual tape loops. The concept is that while each part is repeated, the repeated parts move out of phase with each other and so create new sounds, rhythms and harmonies. No two performances would be the same, since the number of repetitions isn't fixed.

## Easy implementation in the DAW

Creating loops within the DAW is easy. Almost ***too*** easy. But often our "default" is to make those loops the same length. So the same ideas cycle round. But if we make the loops unequal lengths then it'll take longer for those ideas to sync up. And if the loop lengths are not even - break out of the 2, 4, 8, 16 bar patterns - then it'll take longer for them to repeat.

## I'm in my prime

If you remember back to your basic maths from school, you may remember that you can break down most numbers into their prime components: 2, 3, 5, 7, 11, 13, 17, 19 etc. What this means for us as Lazy Producers is that if your loop is a prime number of beats / bars long then it'll take MUCH longer for those loops to sync back up. If you have a 4 bar loop and an 8 bar loop then it only takes two repetitions of the 4 bar loop before you hear the same music idea again. If you have a 3 bar loop and a 5 bar loop then it'll take 15 repetitions before you hear the same idea again. This means then that even for two musical ideas we get a lot of variety when you combine the loops even though both loops are repeating fairly often and fairly quickly. When you combine this idea across more than two loops you can get huge amounts of varying material with very little chance of repetition.

## Fibonacci and the Golden Ratio

So primes are nice for preventing repetition in the grand scheme of things, but there are other options. Again, back to early maths and you might remember Fibonacci sequence where each number is the sum of the previous two numbers, starting with 0 and 1. So the first ten numbers are: 0, 1, 1, 2, 3, 5, 8, 13, 21, 34. Notice that there are some prime numbers in there (2, 3, 5, 13) but also some even numbers (8, 34).

If you take the Fibonacci sequence and layout squares with their sides the length of the Fibonacci number, you get what is known as the "Golden Ratio" where the sizes of the squares are grow in relation to the ratio of the previous two parts. There is a familiar pattern also if you connect arcs - quarter of a circle with radius equal to the size of the square. We often see this pattern in nature - in the pattern of seeds or petals on a plant.

```{r, fig.cap="By Romain - Own work, CC BY-SA 4.0, https://commons.wikimedia.org/w/index.php?curid=114415511"}
knitr::include_graphics(path = file.path(here::here(), "images", "Fibonacci_Spiral.svg"))

```

I'm not going to pretend that it has mystical properties (it's only maths after all) but the combination of primes, even and multiples in the Fibonacci sequence might be the ultimate sweet spot of combining things that occasionally repeat and things that take a long time to repeat.

Suffice to say, it's worth playing with some of these ideas to break out of the predictable loop and make something that perhaps could retain interest for a long period. In our next Recipe section we're going to tap into this idea.
