# Tools - Live 12 Sequencers

Live Suite has had Max for Live sequencer devices for some time, but with Live
12.1 some of those have now been included in a new Sequencers pack. The
sequencers in this pack allow you to specify a pattern of notes, octaves,
velocities, probabilities, and many other attributes and sequence these in ways
that are much more like modular system sequencers, where pattern lengths may
differ from number of beats in a bar, or where you can randomise aspects on the
fly.

```{r}
#| echo: FALSE
knitr::opts_chunk$set(echo = FALSE)
```

::: callout-tip
## Key idea

Why put sequencers in a DAW (which is essentially a sequencer)? Well, because
then you can specify notes you want to play, but have these play in a nonlinear
way, or not follow the overall time signature of your Live set. It opens up
possibilities of playing chosen notes but vary the order these are played
resulting in patterns that vary but in very interesting, evolving and generative
ways.

One limitation of MIDI is that it defines both the note on and note off
information as well as the pitch. In the modular synthesizer world, pitch 
sequence and gate (whether the note occurs or not) sequence can be compleetely
independent, and most interestingly can run at different rates or have different
lengths to create permutations of patterns. The sequencers in this pack offer
a little bit of the modular experience within Live - where pitch, velocity, 
octave and other attributes can have different sequence lengths. In order to 
get the "gate" sequence with MIDI you can set the velocity to zero to specify that 
a given note should not sound.

::: 

## SQ

The SQ sequencer offers a HUGE range of tweakability to specify note pitch,
octave, velocity, length, time-shift, ratchet and condition and have the
sequence of values for each of these run alongside the pitch sequence or operate
with a different sequence length to each other. So for example you could specify
a pitch sequence, then have the note velocities change and rotate with a shorter
sequence length to add some variety. Click on the "link" icons to the right of
the note attributes (e.g. Velocity) to unlink the sequence for that attribute
from the pitch sequence length. Clicking on the "dice" icon will randomise
values. The "Length" attribute defines the length of the sequence, while the
"Dir" setting defines the sequence play order - the sequences also don't need to
run forward, but can run in reverse, ping-pong back and forth (running forward
then backward then forward), in a "snake" pattern or randomly. You can have the
sequence be scale aware so that notes respect the Global Scale setting. You can
change the starting note of the sequence by playing a different MIDI note. You
can set the clock for the sequence to run at a different rate to the Global BPM
setting in the Live set, or even set the clock from an external device. MIDI
pitch information can be entered through the Pitch lane, or can be entered via a
Step Sequenceer within the MIDI tab.

Note (as above in the Key Idea) that setting velocity to zero will stop the 
MIDI note from sounding, so if you would like to have a "gate" sequence, then
you can specify notes which notes not to trigger.

```{r}
knitr::include_graphics(path = file.path(here::here(), "images", "Live12_SQ.png"))
```

This is a really powerful sequencer then, that can provide endless variation and
interest from fairly simple patterns.

## Rhythmic Steps

Rhythmic Steps is quite similar in some ways to the SQ sequencer, in that we can
define the sequence of certain attributes - chance, velocity, MPE slide, and
roll - and vary the sequence length of each of these if we wish. Rhythmic Steps
is set up to work with Drum Racks (you can see which note is going to be played
from the Drum Rack in the left hand side graphic). The "Alt" attribute is
interesting as it allows you to specify the probability of playing an alternate
sound (in blue in the Drum Rack graphic), so you could mix closed hi-hat (second
from top row) with an open hi-hat sound.

```{r}
knitr::include_graphics(path = file.path(here::here(), "images", "Live12_RhythmicSteps.png"))
```

You can also reset the pattern (using the [R] or [Auto] buttons in the image
above) so that the patterns will start back at step 1 again. The [Auto] setting
allows you to specify a number of bars after which the reset will happen if you
don't want to manually trigger this. The icon to the left of the [R] reset
button shows how the Rhythm Sequencer can be used directly from Ableton's Push
controller.

## Step Arp

Much like the above sequencers, Step Arp allows you to tweak and sequence the
attributes of an arpeggiator and have the sequence of those attributes run in
parallel with the pitch sequence which is governed through the Octave and Notes
button settings or unlinked to sequence the attributes with different lengths.

```{r}
knitr::include_graphics(path = file.path(here::here(), "images", "Live12_StepArp.png"))
```

The bar to the right of the sequence allows you to specify a range in which the
values will be randomised, so here this would affect the range of velocities
that would be randomised.
