# Recipe - Audio clip or Sampler into effects {#Chapter-020-Recipe-Sampler_drone}

```{r}
#| echo: false
knitr::opts_chunk$set(echo = FALSE)
```
  
The Sampler instrument from Ableton Live is often used to slice and dice breaks
and loops, or to produce multi-sampled instruments that allow you to recreate
sounds and playback across the keyboard. But in our context we're going to use
it to simply play back a longer textured sound sample and apply some modulation
to the cutoff, producing a drone sound. We can then throw this through a barrage
of effects - either using Return channels or via individual audio effects with
the sampler output routed into each. We're effectively reusing an earlier idea, but this time we're starting with Sampler as our sound source. You could
alternatively use the audio output from 
[PaulStretch](#Chapter-016-Tools-PaulStretch)
as the drone generator. The
benefit of using Sampler is that we can add MIDI effects to (randomly) change
the pitch of the sample so that it doesn't have to be consistent from one loop
to another. Other sample-based instruments like the Max for Live Granulator will allow you to do
similar.

1.  Create a track with the Sampler instrument and one clip with a single note of 8 or 16 bars.

2.  In Sampler's Filter/Global tab, set the amplitude envelope to have a long attack and a very long release (several seconds each). This will allow the notes to blend and overlap smoothly.

3.  In the Sample Modulation section, create some modulation on Filter, Sample
Offset and other parameters so that the sounds morph and change over time rather
than staying static. You can also modulate the LFO amounts or rates in LFO2 and
LFO3 so that the modulation itself isn't static. Modulating the Sample Offset
makes Sample work a little like a granulator - instead of starting at the
beginning of the sample each time, the modulation picks a new position to start
from. This prevents the sound from being too predictable with each new note.

```{r}
#| fig-alt: "Two images showing the Sampler instrument. In front of the Sampler are the Random, Scale and Note Length MIDI devices. The Sampler has a notch filter applied and the frequency of this is being modulated by the LFO. The filter has a slow attack (2 miliseconds), slow delay (10 seconds), slow release (2 seconds). The Envelope of the VCA has slow attack (5 seconds), long release (19 seconds) and maximum sustain level. In the second image we see that LFO 1 - triangle - is modulating Filter and Pan while LFO 2 - Sine - is modulating the amount that LFO 1 is affecting the Pan and Filter. Both LFOs are very slow. LFO 3 which is a Sample and Hold type is modulating the Sample Offset - the starting point of the sample and LFO 2 Amount."
    knitr::include_graphics(path = file.path(here::here(), "images", "Sampler_Drone_Global.png"))
    knitr::include_graphics(path = file.path(here::here(), "images", "Sampler_Drone_Modulation.png"))
```

4.  Using techniques described in 
[Tools - Creative use of audio effects](#Chapter-015-Tools-Audio_Effects)
 and used in the
[Recipe - Making a drone using some audio generators and effects](#Chapter-017-Recipe-Audio_generators_drone)
 set up a network of audio effects on effect
sends / Return channels.

In the example below the Sampler instrument is in Channel 1 and is sending audio
out to a variety of FX channels. Note that the Return channels are also routing
audio around between them to further add layers of effects. You can see that Return A (the "Nice Reverse
+ Reverb" effect) is sending audio out to Return D (Spectral Resonator) and that
that channel is sending audio back to Return A. Using feedback to route audio signal between Return channels is a little risky as sound can sometimes build up quickly if channels are eventually routed back into themselves. Adding a Limiter to the Return or Main channels would prevent runaway sound feedback causing physical damage to ears or speakers.

Spectral Time and Spectral Resonators are
interesting effects to add texture and resonance into fairly static inputs (like
our drone here).

```{r}
#| fig-alt: "The image shows the tracks of the Ableton set. As discussed in the text the Sample audio is being routed to various Return channels which in turn are being routed between each other to create a complex audio effect chain."
knitr::include_graphics(path = file.path(here::here(), "images", "Recipe-Sampler_drone.png"))
```

5.  You can apply MIDI effects to the Sampler track to pick a new random &
quantised note each time the clip restarts. Unlike using an audio clip, using
Sampler allows us to repitch the sound through different MIDI notes.

6.  Once you have a setup that produces pleasing results, try swapping in a different pad sample. Doing this allows you to change up the
sounds generated, as each sample will have different tones, resonances and
features.

7.  Press play. Listen to what you hear, and refine settings if necessary.

------------------------------------------------------------------------

You can hear my version of this track here: 
[https://soundcloud.com/mikeksmith/6_sampler-drone-6](https://soundcloud.com/mikeksmith/6_sampler-drone-6){.uri}

```{r}
#| results: asis
htmltools::includeHTML(path = file.path("HTML","Recipe6-SamplerDrone.html"))
```
