---
editor: 
  markdown: 
    wrap: 72
---

# Recipe - Doing more with less using MIDI effects {#Chapter-006-Recipe-MIDI_tools}

```{r}
#| echo: FALSE
knitr::opts_chunk$set(echo = FALSE)
```

::: callout-tip
## Key idea

In this recipe we'll use the simplest possible MIDI source, a single
MIDI note, and process it with MIDI effects and routing to create an
entire track from one note. We're going to use MIDI effects [Tools -
MIDI tools](#Chapter-004-Tools-MIDI_tools) . We're also going to be
using MIDI and audio routing which is covered in [Process - Routing MIDI
and Audio](#Chapter-005-Process-Routing) . We will also use modulation
in sound design to keep the listener engaged - [Tools -
Modulators](#Chapter-007-Tools-Modulators) .

We're going to have a single MIDI generator track, and then additional
tracks with MIDI Effects to process the inputs. We're going to route the
output of these processed MIDI tracks into various instruments.

You may need to try various sounds and jump between tweaking the MIDI
processing and the sound design in the instruments in order to get the
sound you want.
:::

1.  Create a MIDI track. This is going to be the "generator" for our
    track. We're not going to add an instrument to this track.

2.  Create a single MIDI clip with a single C3 note lasting one bar.
    Name this the "Generator" track (CTRL+R to rename).

3.  Create a new MIDI track called "Note length" and another called
    "Arp". In each of these tracks, set the MIDI input to the Generator
    Track (the single note C3 track) and set Monitor to "In".

```{r}
#| fig-alt: "Image shows the MIDI routing from a generator track, which is creating a single MIDI note of C3 on the first beat of every bar, being sent through Note Length and Arp tracks which are in turn transforming this initial MIDI note."
knitr::include_graphics(path = file.path(here::here(), "images", "Recipe2_MIDI_manipulation.png"))
```

In the "Note Length" track, add a MIDI Note Length MIDI effect. This
will be used to take the output from the Operator track and
stretch out the MIDI notes to last a nominated number of seconds.

```{r}
#| fig-alt: "Image shows the MIDI Note Length device. Here the Length of the MIDI note is extended to 5 seconds."
knitr::include_graphics(path = file.path(here::here(), "images", "Recipe2_MIDI_NoteLength.png"))
```

In the "Arp" track we're going to add an Arpeggiator effect (surprise!)
but also another MIDI "Note Length" effect with which we can tweak the
note length of the output from the Arpeggiator to taste, and a Velocity
effect to add a little variation to the velocity of the output MIDI
notes.

```{r}
#| fig-alt: "Image shows the MIDI Arpeggiator device which is playing notes in the order they appear at a rate of eighth note divisions. Notes are arpeggiating up 7 semitones each time. A Note Length MIDI device comes after this which is extending the notes to 775 miliseconds. Finally a MIDI Velocity device is adding a little random velocity and driving the velocity values upwards slightly."
knitr::include_graphics(path = file.path(here::here(), "images", "Recipe2_MIDI_Arp.png"))
```

Tweaking the Length and Gate settings of "Note Length" will allow us to
alter the sound of what is generated from the "Arpeggiator" effect -
note that the Gate setting in the "Arpeggiator" does this as well, but
we could potentially map these parameters to a MIDI controller if we
wanted to vary the arpeggiator sound coming from this track.

4.  Now let's go on to the instruments and sound design.
    
```{r}
#| fig-alt: "Image shows the various routing paths in the Recipe 2 example and modulation of track mix volume and sends."
knitr::include_graphics(path = file.path(here::here(), "images", "Recipe2_MIDI-and-audio-routing.png"))
```
I have an Operator track with a short, plucky sound. A Wavetable instrument provides a pad sound. An Electric instrument provides an electric piano sound. Finally there's a Bass track providing anchoring bass notes to the track.

5.  I've added a Shaper to the Wavetable track and mapped its output to
    the Gain knob on a Utility device. The "minimum" is set to 50% and
    the "maximum" to 0% and the rate is set to 1/4 quarter notes so that
    each beat, the Shaper ducks the gain to produce a pumping "sidechain
    compressor" like effect.

```{r}
#| fig-alt: "Image shows the Shaper modulation device. The device has a mapping button at the top left which determines which parameter is going to be modulated. The minimum value is set to 50%, the maximum is set to 0% which means that as the modulation goes up, the parameter value will be turned down. A variety of modulation shapes are include along the bottom. There is a representation of the modulation as it repeats in the top right. Rate and depth dials are below this. In the example shown Shaper is being used to turn down the gain of a Utility device."
knitr::include_graphics(path = file.path(here::here(), "images", "Recipe2_MODULATION_Shaper-gain.png"))
```

6.  The second Wavetable track, named “Wavetable pre-compressor,” pulls
    its audio from the main Wavetable track before the ducking effect
    ("Pre-FX"). This track's output is set to "Sends-Only," so we only
    hear its processed signal through the return effects, never the dry
    sound.

7.  On the Operator track I've added additional MIDI Effects to transform the incoming single MIDI note to create movement and interest.

```{r}
#| fig-alt: "Image shows the MIDI device chain which is adding Random MIDI notes, applying a Minor pentatonic scale, using MIDI Note Echo device, another Random MIDI device, another MIDI Note Echo and a final Minor Pentatonic scale device."
knitr::include_graphics(path = file.path(here::here(), "images", "Recipe2_MIDI_fx.png"))
```

You can employ all kinds of MIDI effects here. In this example I have
grouped these effects into a MIDI Effect Rack and I'm using Macros to
allow me to switch on and change various effect parameters from a single
set of 8 knobs.

The "Random" effect is going to add (or subtract) up to 12 semitones
shift to the incoming note. Remember that our generator is a single C3
note for one bar. We then quantise this randomness to the C minor
pentatonic scale using the "Scale" effect.

Then we have two distinct "Note Echo" effects which are going to echo
the incoming note at 5 beats and 8 beats delay. Each effect has the
"Thru" setting switched on so we'll hear the original note and its
delay. You can experiment with setting this to "Mute" to mute the
incoming note, and just play the delayed version. The device on/off
switch on the top left of the device is mapped to a macro knob which
allows me to switch these devices on and off with a knob or button on my
MIDI controller.

The second "Random" effect adds more randomness to the original note,
and all of the delayed / echo output MIDI notes. This allows me to add a
little more variation to the output by turning up the "Chance" setting
on this device. Note that the "Choices" knob is also mapped to a macro
knob, so I can dial up more or less variation.

A third "Note Echo" has a 3 beat delay with a Pitch shift up 7 semitones
but with very low Feedback setting so that the pitching up fades
quickly. Again, device on/off is mapped to a controller knob or button
to allow me to turn this on and off.

Finally a "Scale" device ensures that all this chaos is quantised to a C
minor scale.

These are just examples of what you could do, but the concept here is
that the generating device can be as simple as you like - a single MIDI
note lasting one bar in this case - but these devices then allow you to
create a cascade of evolving MIDI note information all within the C
minor pentatonic scale. Mapping to MIDI controllers allows me to dial up
and back the amount of variation during the track.

8.  Note that I'm routing the output from the Operator track with all the MIDI Effects into the Note Length MIDI track, and the output of that becomes the input of the Arp MIDI track. By routing the MIDI information around like this we ensure that the MIDI information is linked and related giving a more consistent sound.

In performance of this track, I use a MIDI Controller to change settings
of the MIDI Effect Rack in the MIDI note track, which essentially is
using me (the human in charge) to dictate HOW MUCH generative randomness
happens, but the MIDI effects in the "Note Length" and "Arp" tracks
ingest this chaos and make some things happen downstream across a
variety of instruments and sounds. I then use the controller to fade up
and down the volume of each of the instrument tracks to allow me to
shape the overall emerging track so that all sounds aren't playing all
the time.

9.  We have mostly been looking at MIDI ***NOTE*** effects here. There’s
    also a much-overlooked MIDI device called “Expression Control” which
    allows you to map incoming MIDI data (like velocity or pitch bend)
    to almost any parameter in Live, including controls in the
    Operator instrument just for illustration.

```{r}
#| fig-alt: "Image shows the Live 11 Expression control device. In this device various MIDI inputs like Velocity, Pitchbend and Random values are modulation parameters like filter cutoff, synthesizer release time and filter resonance."
knitr::include_graphics(path = file.path(here::here(), "images", "Expression_control.png"))
```

Sure, instruments like Operator and Wavetable have the ability to vary
many different parameters using MIDI information, but notice how each of
the MIDI parameters on the left hand side has a drop-down menu option.
Let's look at what we can do...

```{r}
#| fig-alt: "Image shows that the Expression control device can take a wide range of input controls, which then can modulate various parameters."
knitr::include_graphics(path = file.path(here::here(), "images", "Expression_control_2.png"))
```

There's a wide variety of incoming MIDI inputs which can then be used to
map to effect controls, as well as a "random" input which will select a
new value for every MIDI input note, and "incremental" which will
increment values for every MIDI input note.

Also note that the effect allows you to specify Min and Max levels for
each of these - for example the Random input is mapped to Tone in
Operator, but the range is limited between 55% and 75% of the values.
This kind of constraint can add nice variations in timbre for each note
without it sounding too jarring.

Using MIDI "Expression Control" can be useful to modify a variety of
parameters in Ableton including some that aren't easily mapped within an
instrument.

10. The basic concept in this recipe - and one that I use frequently -
    is to take one MIDI part and make the most of it. Using MIDI effects
    like this can ensure that the resulting sounds are ***related*** to
    the input, without having to be ***exactly*** the same. Anything you
    can do to delay, alter, spread out, harmonise and alter the incoming
    MIDI will add interest to the part. Using these MIDI parts with
    different instrument parts / sounds and audio effects will add
    variety and interest.
    
    
It ***might*** be better here to put the MIDI Effects from the Operator track onto the MIDI Generator track. This would then separate MIDI generation, effect and transform processes from the instrument and sound-design aspects. 

11. Press play. Listen to what you hear, and refine settings if
    necessary.

------------------------------------------------------------------------

You can hear an example of this track here:
[https://soundcloud.com/mikeksmith/2_single-note-input/s-ijkBeXue1aw](https://soundcloud.com/mikeksmith/2_single-note-input/s-ijkBeXue1aw?in=mikeksmith/sets/the-lazy-producer-recipes/s-PGMEWqfwKGz&si=4cb32f77057549d28894966ca45a2715&utm_source=clipboard&utm_medium=text&utm_campaign=social_sharing){.uri}

```{r}
#| results: asis
htmltools::includeHTML(path = file.path("HTML","Recipe2-Single_Note_Input.html"))
```
