# Process - Unsynced loops {#Chapter-021-Process-Unsynced_loops}

```{r}
#| echo: FALSE
knitr::opts_chunk$set(echo = FALSE)
```

> "Around the world" - Daft Punk

Everyone loves a good loop. If you get a good idea of a pattern or
something that lasts 8 bars then you can let that good thing just loop
around and around and around (the world). A large part of The Lazy
Producer's arsenal depends on loops and repetition - after all we're
balancing predictable and unexpected, right? If a voice in your head is saying, “Wait, infinitely repeating 8-bar loops is the ultimate in Lazy Producing,” I’d have to agree. So how do we break away from the tyranny of the 8 bar loop?
Arrangements? Verse-Chorus-Middle 8 structures?

Before you abandon the ideas of generative music (which doesn't really
lend itself well to arrangements, verses or choruses), there's a trick
up our sleeve which is to take our loops and make one of them shorter or
longer than the other.

::: callout-tip
## Key idea

Making loops different lengths combines the predictable: each loop
contains the same music idea each time - with the unexpected: the
combination of those musical ideas each time the loop cycles creates
something new and unheard.

::: 

## Set phases to stunning

Minimalist composers like Steve Reich often worked with musical phrases that were repeated by each performer. By having phrases that were slightly different lengths, they would gradually go out of sync, but then wind up aligned at a different point in the phrase. This method coined the term "phasing" composition.

Sometimes the number of repetitions of each phrase was at the discretion of the 
performer, sometimes governed
by the number of times the performer could play it in one breath,
sometimes the composer worked with actual tape loops of different lengths. The concept is
that while each part is repeated, the repeated parts move out of phase
with each other and so create new sounds, rhythms and harmonies. No two
performances would be the same, since the number of repetitions isn't
fixed.

## Easy implementation in the DAW

Creating loops within the DAW is easy. Almost ***too*** easy. But often
our "default" is to make those loops the same length. So the same ideas
cycle round. But if we make the loops unequal lengths then it'll take
longer for those ideas to sync up. And if the loop lengths are not
even - break out of the 2, 4, 8, 16 bar patterns - then it'll take
longer for them to repeat.

We actually used loops of different lengths in 
[Recipe - Changing chords](#Chapter-001-Recipe-Changing_Chords).

## I'm in my prime

If you recall basic maths from school, you'll remember
that you can break down most numbers into prime factors: 2, 3,
5, 7, 11, 13, 17, 19 etc. What this means for us as Lazy Producers is
that if your loop is a prime number of beats / bars long then it'll take
MUCH longer for those loops to sync back up. If you have a 4 bar loop
and an 8 bar loop then it only takes two repetitions of the 4 bar loop
before you hear the same music idea again. If you have a 3 bar loop and
a 5 bar loop then it'll take 15 repetitions before you hear the same
idea again. This means then that even for two musical ideas we get a lot
of variety when you combine the loops even though both loops are
repeating fairly often and fairly quickly. When you combine this idea
across more than two loops you can get huge amounts of varying material
with very little chance of repetition.

## Fibonacci and the Golden Ratio

Primes are nice for preventing repetition in the grand scheme of
things, but there are other options. Again, back to early maths and you
might remember Fibonacci sequence where each number is the sum of the
previous two numbers, starting with 0 and 1. So the first ten numbers
are: 0, 1, 1, 2, 3, 5, 8, 13, 21, 34. Notice that there are some prime
numbers in there (2, 3, 5, 13) but also some even numbers (8, 34).

If you take the Fibonacci sequence and layout squares with their sides
the length of the Fibonacci number, you get what is known as the "Golden
Ratio" where the sizes of the squares grow in relation to the ratio
of the previous two parts. There is a familiar pattern also if you
connect arcs - quarter of a circle with radius equal to the size of the
square. We often see this pattern in nature - in the pattern of seeds or
petals on a plant.

```{r}
#| fig-cap: "By Romain - Own work, CC BY-SA 4.0, https://commons.wikimedia.org/w/index.php?curid=114415511"
#| fig-alt: "The image shows the fibonacci spiral."
knitr::include_graphics(path = file.path(here::here(), "images", "Fibonacci_Spiral.png"))
```

I won't pretend it has mystical properties (it's just math, after all), but the mix of prime, even, and composite numbers in the Fibonacci sequence might be the perfect sweet spot of combining
things that occasionally repeat and things that take a long time to
repeat.

Suffice to say, it's worth playing with some of these ideas to break out
of the predictable loop and make something that perhaps could retain
interest for a long period. In our next Recipe section we're going to
tap into this idea.

## Let's ROTATE

In discussion of Euclidean Sequencers in 
[Tools - MIDI Generators](#Chapter-012-Tools-MIDI_Generators) 
we talked about how you can rotate a sequence so that it doesn't start at 12 o'clock position (or on the first beat of the sequence).  The same is true of loops. To break out of the predictable sequence you could choose to jump the loop start point to a different position. Sometimes small things like this will unveil new ideas, new rhythms which may be sufficient to keep interest.
