# Recipe - Using the Fibonacci sequence to create loops (after DivKid) {#Chapter-022-Recipe-Fibonacci_loops}

```{r}
#| echo: FALSE
knitr::opts_chunk$set(echo = FALSE)
```
  
This Recipe is adapted from an idea presented by DivKid (Ben Wilson) on his
[Instagram](https://www.instagram.com/divkidvideo) channel. I'm very grateful to
Ben for giving me permission to share this idea in this Recipe. I would
encourage you to check out DivKid's content
(<https://www.youtube.com/divkidvideo>) - he has a lot of really great
tutorials, walkthroughs and ideas for use with modular synth rigs, but which can
also be applied to Ableton Live. You can also sign up to his
[Patreon](https://www.patreon.com/DivKid) where if you sign up as a Patron you
can download a copy of the Ableton set that goes with this Recipe.

1\. Create one loop per track in an Ableton Live set according to the Fibonacci
sequence length. Initially, set up the loops where 1 = 1/16th Note, so 2 =
2/16ths, etc. You can experiment with longer loop lengths where 1 = 1/8th Note,
but obviously these will repeat much more slowly.

2\. Choose the notes and sounds for each loop. You might want to use a combination of drum hits and melodic tones. You can
choose notes from a scale or from a chord e.g. C, Eb, G, Bb, D.

```{r}
#| fig-alt: "The image shows an Ableton Live set where each track has a single MIDI note clip, but each clip has lengths defined by the Fibonacci Series."
knitr::include_graphics(path = file.path(here::here(), "images", "Fibonacci_Polymeters.png"))
```

3\. You can add modulation for the sounds, add randomness in velocity of the
notes to prevent them from becoming too "samey" or you could add the occasional
random / quantised pitch shift.

4\. In his example, DivKid uses plucky sine waves from Operator, but breathes life into the result by adding tone shaping, sound design, and granular delay to the group track.

```{r}
#| fig-alt: "The image shows the effect chain which comprises of Phaser, Saturator, Shifter, Corpus, Grain Delay, Erosion and Multiband compressor in OTT preset."
knitr::include_graphics(path = file.path(here::here(), "images", "Fibonacci_FX.png"))
```

5\. Given this basic set up you could apply many of the concepts and ideas
presented in earlier chapters to further use MIDI in other instruments e.g.
pads, or take the resulting audio and process it further.

6\. Press play. Listen to what you hear, and refine settings if necessary.

Here is DivKid's version of this recipe: <https://www.instagram.com/reel/C2ID25WqB0l/>
